Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-10-08 content.tex
Okay, better fix for soundness issue.

%% FIXME: remove stores from memo-table since they will always be unchanged
%% TODO: Show that store widening and store/memo/KTab widening have a bisimulation.
{
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{4pt}
\setlength{\abovedisplayshortskip}{0pt}
\setlength{\belowdisplayshortskip}{8pt}
\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Store \times \Kont \\
  \System &= \wp(\sa{State}) \times \wp(\sa{State}) \times \KTab \times \Memo \\
  {\mathcal F} &: \sa{\State} \to \System \to \System \\
  {\mathcal F}(\mastate_0)(S, F, \mktab, \mmemo) &= (S \cup S', F', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mktab,\mmemo) \stepto \mstate'} \\
                \wn(\tpl{(\mexpr,\menv),\mstore,\mkont},\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}
\begin{minipage}{.55\linewidth}
  \begin{align*}
    \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
    \mmemo' &= \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I}
  \end{align*}
\end{minipage}
\begin{minipage}{.40\linewidth}
  \begin{align*}
    S' &= \setbuild{\mastate'}{\wn(\mastate', \_, \_, \_) \in I} \cup \set{\mastate_0}\\
    F' &= S' \setminus S
  \end{align*}
\end{minipage}
}
%%

@@

%% TODO (thanks to ccshan): Discuss difference between CPS for shift/reset and this approach,
%% Discuss how this does not generalize to first class prompts due to append versus cons.

@@

\begin{figure*}
  \centering
% Shift
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mmktab, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \mstore',\kmt,\mmkont, \mmktab', \mktab,\mmemo}$
    \\
    where & $\mstore' = \joinone{\mstore}{\maddr}{\vcomp{\mkont'}}$ \\
    & $(\mkont',\mmktab') = \approximate(\mkont, \mmktab, \maddr)$
\\
% Return
   $\tpl{\mval, \mstore, \krt{\mctx}, \mmkont, \mmktab, \mktab, \mmemo}$
   &
   $\tpl{\mval, \mstore, \mkont, \mmkont, \mmktab, \mktab, \mmemo'}$
   if $\mkont \in \returns(\mmktab, \mktab, \mctx)$
   \\
   where & $\mmemo' = \mmemo \sqcup \memoize(\mmktab, \mctx, (\mval, \mstore, \mmktab))$
  \end{tabular}
  \caption{Fixed table-based semantics for shift/reset}
  \label{fig:shift-reset-table1}
\end{figure*}

The meta-functions mentioned in the fixed semantics all deal with the addition of $\maddr$ to contexts.
%
If the context in an $\rt$ continuation is approximate, we must return to all the continuations known for all the contexts it approximates:
\begin{align*}
  \returns(\mmktab, \mktab, ((\mexpr,\menv), \mstore)) &= \mktab((\mexpr,\menv), \mstore) \\
  \returns(\mmktab, \mktab, ((\mexpr,\menv), \maddr)) &=
    \bigcup\setbuild{\mktab((\mexpr,\menv),\mstore)}{\mstore \in \mmktab(\maddr)}
\end{align*}

At return boundaries, the memo table must add the result to all the represented contexts:
\begin{align*}
  \memoize(\mmktab, ((\mexpr,\menv), \mstore'), \mathit{out}) &=
  [((\mexpr,\menv), \mstore') \mapsto \set{{\mathit{out}}}] \\
  \memoize(\mmktab, ((\mexpr,\menv), \maddr), \mathit{out}) &=
  \bigsqcup\limits_{\mstore' \in \mmktab(\maddr)}\snglm{((\mexpr,\menv), \mstore')}{\mathit{out}}
\end{align*}
At capture time, we strip the store in the $\rt$ continuation if there is one, and replace it with an address to make the context storeable:

\newcommand{\replacectx}{\mathit{replace}\mstore}
\newcommand{\addstore}{\mathit{add}\mstore}
\begin{align*}
  \approximate(\mkont, \mmktab, \maddr) &= (\replacectx(\mkont), \addstore(\mkont)) \\
  \text{where }
   \replacectx(\kmt) &= \kmt \\
   \replacectx(\krt{((\mexpr,\menv),\_)}) &= \krt{((\mexpr,\menv),\maddr)} \\
   \replacectx(\kcons{\mkframe}{\mkont}) &= \kcons{\mkframe}{\replacectx(\mkont)}
  \\[2pt]
   \addstore(\kmt) &= \mmktab \\
   \addstore(\krt{((\mexpr,\menv),\mstore)}) &= \joinone{\mmktab}{\maddr}{\mstore} \\
   \addstore(\krt{((\mexpr,\menv),\maddr')}) &= \joinm{\mmktab}{\maddr}{\mmktab(\maddr')} \\
   \addstore(\kcons{\mkframe}{\mkont}) &= \addstore(\mkont)
\end{align*}

@@

We prove this with the invariant that the continuation and meta-continuation have unique unrollings in addition to the table invariants from \autoref{sec:pdcfa}, since entries in $\mktab$ are always fresh.

@@

%% TODO: complete
\paragraph{Comparison to PDCFA CPS'd to remove {\tt shift} and {\tt reset}:}{
We lose precision if we use a CPS transform to compile away {\tt shift} and {\tt reset} forms, because variables are treated less precisely than continuations.
%
In pushdown analysis, we track function return points with entire calling contexts rather than the $k$-CFA way of allocating the continuation in the heap at some lower-precision address, like a variable.
%
In our summarizing analysis of delimited composable control, we get the same added precision for {\tt reset}, {\tt shift}, and continuation invocation points.
%
Consider the following program and its CPS transform for comparison:
% \begin{lstlisting}[mathescape]
% (let* ([id ($\lambda$ (x) x)]
%        [f ($\lambda$ (y) (shift k (k (k y))))]
%        [g ($\lambda$ (z) (reset (id (f z))))])
%   (<= (g 0) (g 1)))
% \end{lstlisting}
\begin{SCodeFlow}\begin{RktBlk}\begin{SingleColumn}\RktPn{(}\RktSym{let*}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktPn{[}\RktSym{id}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{$\lambda$}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{x}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{x}\RktPn{)}\RktPn{]}

>>>>

{
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{4pt}
\setlength{\abovedisplayshortskip}{0pt}
\setlength{\belowdisplayshortskip}{8pt}
\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Store \times \Kont \\
  \System &= \wp(\sa{State}) \times \wp(\sa{State}) \times \KTab \times \Memo \\
  {\mathcal F} &: \sa{\State} \to \System \to \System \\
  {\mathcal F}(\mastate_0)(S, F, \mktab, \mmemo) &= (S \cup S', F', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mktab,\mmemo) \stepto \mstate'} \\
                \wn(\tpl{(\mexpr,\menv),\mstore,\mkont},\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}
\begin{minipage}{.55\linewidth}
  \begin{align*}
    \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
    \mmemo' &= \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I}
  \end{align*}
\end{minipage}
\begin{minipage}{.40\linewidth}
  \begin{align*}
    S' &= \setbuild{\mastate'}{\wn(\mastate', \_, \_, \_) \in I} \cup \set{\mastate_0}\\
    F' &= S' \setminus S
  \end{align*}
\end{minipage}
}
%%

@@

%% (thanks to ccshan): Discuss difference between CPS for shift/reset and this approach,
%% Discuss how this does not generalize to first class prompts due to append versus cons.

@@

\begin{figure*}
  \centering
% Shift
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mmktab, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \mstore',\kmt,\mmkont, \mmktab', \mktab,\mmemo}$
    \\
    where & $\mstore' = \joinone{\mstore}{\maddr}{\vcomp{\mkont'}}$ \\
    & $(\mkont',\mmktab') = \approximate(\mkont, \mmktab, \maddr)$
\\
% Return
   $\tpl{\mval, \mstore, \krt{\mctx}, \mmkont, \mmktab, \mktab, \mmemo}$
   &
   $\tpl{\mval, \mstore, \mkont, \mmkont, \mmktab, \mktab, \mmemo'}$
   if $\mkont \in \returns(\mmktab, \mktab, \mctx)$
   \\
   where & $\mmemo' = \mmemo \sqcup \memoize(\mmktab, \mktab, \mctx, (\mval, \mstore, \mmktab))$
  \end{tabular}
  \caption{Fixed table-based semantics for shift/reset}
  \label{fig:shift-reset-table1}
\end{figure*}

The meta-functions mentioned in the fixed semantics deal with the addition of $\maddr$ to contexts.
%
At capture time, we strip the store and stored continuations closure in the $\rt$ continuation if there is one, and replace them with an address to make the context storeable.
%
We can't recursively map addresses to stores and continuation closures, so we flatten the context's closure into the current one.
%
This operation makes lookups more costly since we have to find closures bounded above by the current closure, rather than one exact closure.

\newcommand{\replacectx}{\mathit{replace}\mstore}
\newcommand{\addstore}{\mathit{add}\mstore}
\begin{align*}
  \approximate(\mkont, \mmktab, \maddr) &= (\replacectx(\mkont), \addstore(\mkont)) \\
  \text{where }
   \replacectx(\kmt) &= \kmt \\
   \replacectx(\krt{((\mexpr,\menv),\ldots)}) &= \krt{((\mexpr,\menv),\maddr)} \\
   \replacectx(\kcons{\mkframe}{\mkont}) &= \kcons{\mkframe}{\replacectx(\mkont)}
  \\[2pt]
   \addstore(\kmt) &= \mmktab \\
   \addstore(\krt{((\mexpr,\menv),\mstore,\mmktab')}) &= \joinone{\mmktab}{\maddr}{\mstore}\sqcup \mmktab' \\
   \addstore(\krt{((\mexpr,\menv),\maddr')}) &= \joinm{\mmktab}{\maddr}{\mmktab(\maddr')} \\
   \addstore(\kcons{\mkframe}{\mkont}) &= \addstore(\mkont)
\end{align*}
%
If the context in an $\rt$ continuation is approximate, we must return to all the continuations known for all the contexts it approximates:
\begin{align*}
  \returns(\mmktab, \mktab, ((\mexpr,\menv), \mstore, \mmktab')) &= \mktab((\mexpr,\menv), \mstore, \mmktab') \\
  \returns(\mmktab, \mktab, ((\mexpr,\menv), \maddr)) &=
    \bigcup\setbuild{\mktab((\mexpr,\menv),\mstore,\mmktab')}{(\mstore \in \mmktab(\maddr), \mmktab' \sqsubseteq \mmktab}
\end{align*}

At return boundaries, the memo table must add the result to all the represented contexts:
\begin{align*}
  \memoize(\mmktab, \mktab, ((\mexpr,\menv), \mstore', \mmktab'), \mathit{out}) &=
  [((\mexpr,\menv), \mstore',\mmktab') \mapsto \set{{\mathit{out}}}] \\
  \memoize(\mmktab, \mktab, ((\mexpr,\menv), \maddr), \mathit{out}) &= [\mctx \mapsto \set{\mathit{out}} : \mctx \in \dom(\mktab) \cap C] \\
    \text{where } C &= \setbuild{((\mexpr,\menv), \mstore', \mmktab')}{\mstore' \in \mmktab(\maddr), \mmktab' \sqsubseteq \mmktab}
\end{align*}

@@

We prove this with the invariant that the continuation and meta-continuation have unique unrollings in addition to the table invariants from \autoref{sec:pdcfa}, since the distinct $\mstore$ at push and capture time makes entries in $\mktab$ unique, regardless of $\mmktab$.

@@

\paragraph{Comparison to PDCFA CPS'd to remove {\tt shift} and {\tt reset}:}{
We lose precision if we use a CPS transform to compile away {\tt shift} and {\tt reset} forms, because variables are treated less precisely than continuations.
%
In pushdown analysis, we track function return points with entire calling contexts rather than the $k$-CFA way of allocating the continuation in the heap at some lower-precision address, like a variable.
%
In our summarizing analysis of delimited composable control, we get the same added precision for {\tt reset}, {\tt shift}, and continuation invocation points.
%
Consider the following program and its CPS transform for comparison:
% \begin{lstlisting}[mathescape]
% (let* ([id ($\lambda$ (x) x)]
%        [f ($\lambda$ (y) (shift k (k (k y))))]
%        [g ($\lambda$ (z) (reset (id (f z))))])
%   (<= (g 0) (g 1)))
% \end{lstlisting}
\begin{SCodeFlow}\begin{RktBlk}\begin{SingleColumn}\RktPn{(}\RktSym{let*}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktPn{[}\RktSym{id}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{$\lambda$}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{x}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{x}\RktPn{)}\RktPn{]}
